Hydraulically-vulnerable trees are buffered by deep-water access in a tropical forest community

### 2020-10-10 manuscript.rmd
growth vs. dbh relationship without color coding by census interval
1. Added a schematic fig. for the methods workflow 2. Updated extreme drought year fig. description

\
In summary, we obtained daily dynamics of plant water potential by depth in the whole-soil column with a locally parameterized and calibrated 1-D hydrological water balance model. We developed a novel inverse model to estimate species effective rooting depths (ERD), defined as the depth at which soil water dynamics best explained observed dynamics of species growth. This model incorporated the impact of atmospheric drought on growth and was constrained by above-ground hydraulic traits, namely, species leaf vulnerability curves. For the latter, we collected new data on vulnerability curves for 21 species and scale-up to the community via scaling relationships with 'softer' traits, namely, Wood Density and Leaf Mass Area. We evaluated ERD against independent estimates based on existing data on plant and soil water isotopes. We then tested whether deeper ERD was associated with hydraulically efficient but vulnerable traits (Table 1 ), but not greater mortality rates.

@@

```{r dbh-model, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', dpi = 96, out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-dbh-model}A scatterplot of tree growth rates (circles) from all census intervals versus dbh at the beginning of the census interval is overlaid with a dbh model of growth (spline). Growth rates are color coded by census interval. The model equation is shown in inset along with goodness-of-fit (Adj. *R*^2^).")}
knitr::include_graphics(file.path("figures/dbh/growthrate_dbh_predicted_growth_with_spline_degree_5_dbh_cutoff_100_model.jpeg"))
```

@@

```{r swp-droughts, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-droughts}Extreme soil droughts by census interval (panels) at three representative depths: 0.5, 1 and 1.7 m, depicted in different colors. Mean soil water potential ($Psi_{soil,~z}$) for a given DOY and depth z are shown as lines. Filled areas show extreme drought periods, when $Psi_{soil,~z}$ declined below 2.5 percentile of $Psi_{soil,~z}$ for the given DOY and depth z in any given year within the census intervals. Extreme drought periods from multiple years may be overlaid within a panel. Wet season is shown in gray beginning with DOY 120."}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi_model_daily_bestfit_params.top.few_CI_full_interval_panels_climatology_over_study_period_q2.5_depths_in_inverse_model.jpeg"))
```

>>>>

\
In summary (Fig. \@ref(fig:schematic)), we obtained daily dynamics of plant water potential by depth in the whole-soil column with a locally parameterized and calibrated 1-D hydrological water balance model. We developed a novel inverse model to estimate species effective rooting depths (ERD), defined as the depth at which soil water dynamics best explained observed dynamics of species growth. This model incorporated the impact of atmospheric drought on growth and was constrained by above-ground hydraulic traits, namely, species leaf vulnerability curves. For the latter, we collected new data on vulnerability curves for 21 species and scale-up to the community via scaling relationships with 'softer' traits, namely, Wood Density and Leaf Mass Area. We evaluated ERD against independent estimates based on existing data on plant and soil water isotopes. We then tested whether deeper ERD was associated with hydraulically efficient but vulnerable traits (Table 1 ), but not greater mortality rates.

```{r schematic, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-schematic}A schematic for the methods workflow.")}
knitr::include_graphics(file.path("figures/schematic.png"))
```

@@

```{r dbh-model, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', dpi = 96, out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-dbh-model}A scatterplot of tree growth rates (circles) from all census intervals versus dbh at the beginning of the census interval is overlaid with a dbh model of growth (spline). Growth rates are color coded by census interval. The model equation is shown in inset along with goodness-of-fit (Adj. *R*^2^).")}
knitr::include_graphics(file.path("figures/dbh/growthrate_dbh_predicted_growth_with_spline_degree_5_dbh_cutoff_100_model_no_color.jpeg"))
```

@@

```{r swp-droughts, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-droughts}Occurrence of extreme soil droughts by census intervals (horizontal panels) at three representative depths (z): 0.1, 0.6 and 1 m (vertical panels). Mean (black lines) and lower half of 95% distribution (grey areas) of soil water potential ($Psi_{soil,~z}$) for a given DOY and depth z are shown (same across census intervals). $Psi_{soil,~z}$ by DOY for a year (colored line) is only shown if for some part of that year $Psi_{soil,~z}$ declined below the 2.5 percentile distribution. Wet season is shown in gray beginning with DOY 120."}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi_model_daily_bestfit_params.top.few_CI_full_interval_panels_climatology_over_study_period_q2.5_by_dpeth.jpeg"))
```
